Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-03-21 manuscript.Rmd
correct palm biomass + add results and discussion

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Marguerite Cogné, Régis Peltier, Vincent Freycon, Alexis Toumazeau, Camille Piponiot"
date: "2024-03-19"
output: 
  bookdown::word_document2:
  fig_caption: yes
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results = "hide", tinytex.clean = FALSE)
packages_needed = c("ggplot2", "readxl", "data.table", "knitr", "BIOMASS", "terra", "utils", "car")

@@

**Word count**: xxx; **Figure and table count**: 5

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastoral systems, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 255 MgC/ha, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

Although AFS can make agriculture more resilient while improving its carbon footprint, poor adaptation to the local context can explain many failures to introduce new agroforestry techniques [@Coe2014]. Conversely, some traditional agroforestry techniques have evolved in a particular context (biophysical, social and economic) and are therefore adapted to it [@Gouyon1993;@Aumeeruddy1994;@Peltier1996]. Moreover, traditional small-scale AFS typically retain high levels of biodiversity, making them key elements for biodiversity conservation in the tropics [@Perfecto2008;@Torquebiau1992]. A better understanding of traditional agroforestry practices and their ecological characteristics could provide better support for the local development of agroforestry.

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods, but food security is threatened by population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime [@WB2009;@Molyneux2012]. In this context, productive, resilient and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. Better quantification of these carbon stocks would improve the understanding of the ecological importance of these AFS at both local and global scales, and help to integrate them into national and international climate change mitigation initiatives.

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Fig. \@ref(fig:map-sites)). The average soil properties differed between sites (Table \@ref(tab:stab-site-soil)), with Gariuai having Clay Loam soil texture compared to Osso Luga having clay soil texture. The two sites also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2000. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

We used the typology described in @Cogne2024, developed in the same study area. This typology was based on the one proposed by @Nair1993;@Nair2022, and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFS identified were: crop system including a wooded fallow phase (CF), Sylvopastoral (SP), Young agroforest (YA), Home garden (HG) and Forest garden (FG). CF are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SP land is a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YA are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years, usually with the help of projects. HG are plots of land close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFS are presented in more detail in the supplemental information (Table \@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected 3 plots from each of the 5 traditional AFS identified in each of the two sites (30 plots in total; Fig. \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06-1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Fig. \@ref(fig:design)): within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were done with a dendrometer.

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using iNaturalist and Pl\@ntnet application, Geoffrey Hull’s monograph [@Hull2006], the data bases of “the useful tropical plants” (https://tropical.theferns.info/) as well as the native plants websites of the NT herbarium from the Australian government (https://nt.gov.au/environment/native-plants/native-plants-and-nt-herbarium). The correspondence between tetum names and scientific names is provided in the supplemental material (Table \@ref(tab:tetum-names)).

@@

In each of the 30 sampling plots, a composite of 3 samples were collected following a transect to cover the plot’s inner-variability (Fig. \@ref(fig:design)). Soils were sampled with an auger at 0-10 cm and 20-30 cm. Soils samples were then air-dried and sieved at 2 mm. They were the subject of standard physical and chemical analyses at Cirad laboratories in Montpellier, France. Among these analyses, total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

@@

@@

### estimate AGC
data_sim_agc <- data.table(
  data_tot,
  BIOMASS::AGBmonteCarlo(
    D = data_tot$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tot$HEIGHT, 
    errH = data_tot$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = paste0("V", 1:1000),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

## height in cm, with a conversion factor between 0.8 and 1
data_sim_agc[GENUS %in% palm_genera, 
             HEIGHT := rnorm(1, HEIGHT, HEIGHT*cvH)*
               runif(1, .8, 1)* 100]

## add WD data
data_wd_palms = unique(data_wd[data_wd$genus %in% palm_genera, c("genus", "meanWD", "sdWD")])
data_sim_agc = merge(data_sim_agc, data_wd_palms, all.x = TRUE, by.x = "GENUS", by.y = "genus")
data_sim_agc[GENUS %in% palm_genera, WD := rnorm(1, meanWD,sdWD)]

data_sim_agc[GENUS %in% palm_genera, 
             AGC := WD * HEIGHT * (DIAM_130_CM/2)^2 * pi * 
               0.89 * # volumetric shrinkage
               1e-6  # convert from g to tons
]
```

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as the volume of a cylinder, using the height and diameter measured in situ, with a volumetric shrinkage of 11% [@WD2023]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height.

@@

# melt data frame
data_sim <- melt(data_sim_agc, id.vars = c("N_PLOT", "N_Q_ID", "ID_VILLAGE", "ID_AF", "iter", "area"), measure.vars = c("AGC", "BGC"))

# estimate biomass per quadrat and plot
data_sim <- data_sim[, .(value = sum(value/area)), .(N_PLOT, N_Q_ID, ID_VILLAGE, ID_AF, iter, variable)]

## bootstrap AGC and BGC values in small plots
data_sim = 
  merge(
    data_sim[!grepl("Q", N_Q_ID)], 
    data_sim[grepl("Q", N_Q_ID), .(
      valueSmall = ifelse(
        length(value) == 0,   ## some quadrats have zero trees: complete their values
        0, 
        mean(sample(c(value, rep(0, 5 - length(value))), replace = TRUE))
      )), .(N_PLOT, variable, iter)], 
    by = c("N_PLOT", "variable", "iter")
  )

data_sim[, value := value + valueSmall]
```

@@

$$SOC_i = \frac{\sum_i Corg_i \cdot (mfine_i \cdot 10-6) \cdot (Splot \cdot \Delta d)}{Vcyl}$$

@@

## error propagation using a Monte Carlo method -------------
data_sim_soil = data_soil[, .(mfine = rnorm(1000, Sample_DRY, Sample_DRY*0.1), # 10% error on fine soil mass
                              Corg = rnorm(1000, Corg, Corg*0.12), # 12% error on C content measurement
                              iter = paste0("V", 1:1000)), 
                          .(N_PLOT, SAMPLE_DEPTH, h1, h0)]

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo() function from the BIOMASS package, using a (rather conservative) coefficient of variation of 15% on height measurements (Hunter et al 2013); large and small errors on 5 and 95% of trees respectively (Chave et al 2004; Réjou-Méchain et al., 2017).
To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (Cirad laboratory in Montpellier, France), 10% for the mass of the fine fraction (Page-Dumroese et al. 1999).

@@

```{r boot-results}
# plot level statistics
dataC_plot = data_sim[, .(
  mean = mean(value),
  lwr = quantile(value, 0.025),
  med = quantile(value, 0.5),
  upr= quantile(value, 0.975)
), .(plot = N_PLOT, site, afs = ID_AF, variable)]

# AFS level statistics (with bootstrapping)
dataC_afs = data_sim[, .(
  mean = mean(value),
  meanBs = mean(sample(value, replace = TRUE))
), .(afs = ID_AF, variable, iter)][, .(
  mean = mean(mean),        
  lwr_mean = quantile(meanBs, 0.025),   
  upr_mean = quantile(meanBs, 0.975)
), .(afs, variable)]

# site level statistics (with bootstrapping)
dataC_site = data_sim[, .(
  mean = mean(value),
  meanBs = mean(sample(value, replace = TRUE))
), .(site, variable, iter)][, .(
  mean = mean(mean),        
  lwr_mean = quantile(meanBs, 0.025),   
  upr_mean = quantile(meanBs, 0.975)
), .(site, variable)]
```

@@

@@

```{r summary, results = "show"}
# calculate median and 95% confidence interval at site level
dataC_afs[, value := paste0(
  signif(mean, 3), " (",
  signif(lwr_mean, 3), "-", 
  signif(upr_mean, 3), ")")] |> 
  subset(variable != "SOC10") |> 
  dcast(afs ~ variable, value.var = "value") |> 
  knitr::kable(col.names = c("Agroforestry system", 
                             "Aboveground carbon (MgC/ha)", 
                             "Belowground carbon (MgC/ha)", 
                             "Soil organic carbon (MgC/ha)", 
                             "Total carbon (MgC/ha)")) 
```

```{r anova, results = "show"}
knitr::kable(anova_table[, c("variable", "factor", "pvalue")])
```

```{r map-sites, fig.height = 6, fig.width = 7}

## site location: create box around plots
data_box = data_plot[, .(xc = mean(long), yc = mean(lat)), .(site)]
data_box[site == "Gariuai", `:=`(w = 0.1, xc = xc + 0.001)]
data_box[site == "Osso Luga", w := 0.05]
data_box[, `:=`(x1 = xc - w/2, x2 = xc + w/2, y1 = yc - w/2, y2 = yc + w/2)]

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Marguerite Cogné, Régis Peltier, Vincent Freycon, Alexis Toumazeau, Camille Piponiot"
date: "2024-03-19"
output: officedown::rdocx_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results = "hide")
packages_needed = c(
  "ggplot2",
  "readxl",
  "data.table",
  "knitr",
  "BIOMASS",
  "terra",
  "utils",
  "car",
  "lmerTest",
  "performance"
)

@@

**Word count**: 7200 (with SI); **Figure and table count**: 5 + 6 in SI

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastoral systems, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

<!-- Although AFS can make agriculture more resilient while improving its carbon footprint, poor adaptation to the local context can explain many failures to introduce new agroforestry techniques [@Coe2014]. Conversely, some traditional agroforestry techniques have evolved in a particular context (biophysical, social and economic) and are therefore adapted to it [@Gouyon1993;@Aumeeruddy1994;@Peltier1996]. Moreover, traditional small-scale AFS typically retain high levels of biodiversity, making them key elements for biodiversity conservation in the tropics [@Perfecto2008;@Torquebiau1992]. A better understanding of traditional agroforestry practices and their ecological characteristics could provide better support for the local development of agroforestry. -->

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods, but food security is threatened by population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime [@WB2009;@Molyneux2012]. In this context, productive, resilient and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024] and can be key elements for biodiversity conservation in the tropics [@Perfecto2008;@Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. Better quantification of these carbon stocks would improve the understanding of the ecological importance of these AFS at both local and global scales, and help to integrate them into national and international climate change mitigation initiatives.

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Figure\ \@ref(fig:map-sites)). The average soil properties differed between sites (Table S\@ref(tab:site-soil)), with Gariuai having Clay Loam soil texture compared to Osso Luga having clay soil texture. The two sites also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2000. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

We used the typology described in @Cogne2024, developed in the same study area. This typology was based on the one proposed by @Nair1993;@Nair2021, and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFS identified were: crop system including a wooded fallow phase (CF), Sylvopastoral (SP), Young agroforest (YA), Home garden (HG) and Forest garden (FG). CF are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SP land is a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YA are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years, usually with the help of projects. HG are plots of land close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFS are presented in more detail in the supplemental information (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected 3 plots from each of the 5 traditional AFS identified in each of the two sites (30 plots in total; Figure\ \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06-1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Figure\ S\@ref(fig:design)): within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were done with a dendrometer.

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using iNaturalist and Pl\@ntnet application, Geoffrey Hull’s monograph [@Hull2006], the data bases of “the useful tropical plants” (https://tropical.theferns.info/) as well as the native plants websites of the NT herbarium from the Australian government (https://nt.gov.au/environment/native-plants/native-plants-and-nt-herbarium). The correspondence between tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

In each of the 30 sampling plots, a composite of 3 samples were collected following a transect to cover the plot’s inner-variability (Figure\ S\@ref(fig:design)). Soils were sampled with an auger at 0-10 cm and 20-30 cm. Soils samples were then air-dried and sieved at 2 mm. They were the subject of standard physical and chemical analyses at Cirad laboratories in Montpellier, France. Among these analyses, total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

@@

# estimate mean AGC
data_tot$mean <- BIOMASS::computeAGB(
  D = data_tot$DIAM_130_CM, 
  WD = data_wd$meanWD, 
  H = data_tot$HEIGHT
) * 0.4713 # conversion factor from biomass to carbon

### estimate AGC with uncertainty propagation

@@

data_sim_agc <- data.table(
  data_tot,
  BIOMASS::AGBmonteCarlo(
    D = data_tot$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tot$HEIGHT, 
    errH = data_tot$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

## height in cm, with a conversion factor between 0.8 and 1
data_sim_agc[GENUS %in% palm_genera, 
             HEIGHT := ifelse(
               iter == "mean", 
               HEIGHT*.9* 100,  ## height in cm
               rnorm(length(HEIGHT), HEIGHT, HEIGHT*cvH)*
                 runif(length(HEIGHT), .8, 1)* 100
             )]

## add WD data
data_wd_palms = unique(data_wd[data_wd$genus %in% palm_genera, c("genus", "meanWD", "sdWD")])
data_sim_agc = merge(
  data_sim_agc,
  data_wd_palms,
  all.x = TRUE,
  by.x = "GENUS",
  by.y = "genus"
)
data_sim_agc[GENUS %in% palm_genera, 
             WD := ifelse(iter == "mean", 
                          meanWD, 
                          rnorm(length(meanWD), meanWD,sdWD))]

data_sim_agc[GENUS %in% palm_genera, 
             AGC := WD * HEIGHT * (DIAM_130_CM/2)^2 * pi * 
               0.89 * # volumetric shrinkage
               1e-6 * # convert from g to tons
               0.4713 # conversion factor from biomass to carbon
]
```

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as the volume of a cylinder, using the height and diameter measured in situ, with a volumetric shrinkage of 11% [@WD2023]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

# melt data frame
data_sim <-
  melt(
    data_sim_agc,
    id.vars = c("N_PLOT", "N_Q_ID", "ID_VILLAGE", "ID_AF", "iter", "area"),
    measure.vars = c("AGC", "BGC")
  )

# estimate biomass per quadrat and plot
data_sim <-
  data_sim[, .(value = sum(value / area)), .(N_PLOT, N_Q_ID, ID_VILLAGE, ID_AF, iter, variable)]

## bootstrap AGC and BGC values in small plots (except for iter = mean)
data_sim = 
  merge(
    data_sim[!grepl("Q", N_Q_ID)], 
    rbind(
      data_sim[grepl("Q", N_Q_ID) & iter != "mean", .(
        valueSmall = ifelse(
          length(value) == 0,   ## some quadrats have zero trees: complete their values
          0, 
          mean(sample(c(value, rep(0, 5 - length(value))), replace = TRUE))
        )), .(N_PLOT, variable, iter)], 
      data_sim[grepl("Q", N_Q_ID) & iter == "mean", 
               .(valueSmall = sum(value / 5)), .(N_PLOT, variable, iter)]
    ),
    by = c("N_PLOT", "variable", "iter"), all = TRUE
  )

# replace NAs is valueSmall (ie no trees in quadrats) with 0
data_sim[, valueSmall := ifelse(is.na(valueSmall), 0, valueSmall)]

## sum mean biomass in quadrats and entire plot
data_sim[, value := value + valueSmall]
```

@@

$$SOC_i = \frac{\sum_i Corg_i \cdot (mfine_i \cdot 10^{-6}) \cdot (Splot \cdot \Delta d)}{Vcyl}$$

@@

## error propagation using a Monte Carlo method -------------
data_sim_soil = data_soil[, .(mfine = c(Sample_DRY, rnorm(1000, Sample_DRY, Sample_DRY*0.1)), # 10% error on fine soil mass
                              Corg = c(Corg, rnorm(1000, Corg, Corg*0.12)), # 12% error on C content measurement
                              iter = c("mean", paste0("V", 1:1000))), 
                          .(N_PLOT, SAMPLE_DEPTH, h1, h0)]

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo() function from the BIOMASS package, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013]; large and small errors on 5 and 95% of trees respectively [@Chave2004;@Rejou-Mechain2017].
To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (Cirad laboratory in Montpellier, France), 10% for the mass of the fine fraction [@Page1999].

@@

```{r boot-results}
# plot level statistics
dataC_plot = data_sim[, .(
  mean = value[iter=="mean"],
  lwr = quantile(value, 0.025),
  upr= quantile(value, 0.975)
), .(plot = N_PLOT, site, afs = ID_AF, variable)]

# AFS level statistics (with bootstrapping)
dataC_afs = data_sim[, .(
  mean = mean(value),
  meanBs = mean(sample(value, replace = TRUE))
), .(afs = ID_AF, variable, iter)][, .(
  mean = mean[iter=="mean"],        
  lwr_mean = quantile(meanBs, 0.025),   
  upr_mean = quantile(meanBs, 0.975)
), .(afs, variable)]

# site level statistics (with bootstrapping)
dataC_site = data_sim[, .(
  mean = mean(value),
  meanBs = mean(sample(value, replace = TRUE))
), .(site, variable, iter)][, .(
  mean = mean[iter=="mean"],    
  lwr_mean = quantile(meanBs, 0.025),   
  upr_mean = quantile(meanBs, 0.975)
), .(site, variable)]

# AFS + site level statistics (with bootstrapping)
dataC_afs_site = data_sim[, .(
  mean = mean(value),
  meanBs = mean(sample(value, replace = TRUE))
), .(afs = ID_AF, site, variable, iter)][, .(
  mean = mean[iter=="mean"],        
  lwr_mean = quantile(meanBs, 0.025),   
  upr_mean = quantile(meanBs, 0.975)
), .(afs, site, variable)]
```

@@

```{r lmereg}
data_reg = melt(dataC_plot, measure.vars = c("mean", "lwr", "upr"), variable.name = "stat") |> 
  dcast(plot+ site+ afs ~ variable + stat)

## mixed effect linear regression
lmeReg <- lmerTest::lmer(log(SOC_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R2 = performance::r2_nakagawa(lmeReg)
summary(lmeReg)

# same for topsoil layer (0-10 cm)
lmeReg10 <- lmerTest::lmer(log(SOC10_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R210 = performance::r2_nakagawa(lmeReg10)
summary(lmeReg10)

dfreg = 
  data.table(
    depth = c("Soil depth: 0-30 cm", "Soil depth: 0-10 cm"), 
    intercept = c(summary(lmeReg)$coefficients[1,1], summary(lmeReg10)$coefficients[1,1]), 
    slope = c(summary(lmeReg)$coefficients[2,1], summary(lmeReg10)$coefficients[2,1]), 
    pvalue = c(summary(lmeReg)$coefficients[2,5], summary(lmeReg10)$coefficients[2,5]),
    R2_cond = c(R2$R2_conditional, R210$R2_conditional), 
    R2_marg = c(R2$R2_marginal, R210$R2_marginal)
  )
```

@@

Total carbon stocks varied between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

AGC stocks varied between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$) than those in Osso Luga (`r dataC_site[site=="Osso Luga" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (yet slightly stronger; p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The proportion of variance explained by the linear mixed models were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths respectively, but a marginal R$^2$ of just `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`% respectively when only the fixed effect was considered (Figure\ \@ref(fig:plot-lmereg)). Overall, AGC stocks were a poor predictor of SOC stocks.

# Discussion

## Large C stocks in traditional AFS in Timor Leste

Compared to values reported in the literature for other AFS, the carbon stocks in this study were particularly high. The value of SOC stocks was similar to other reported values [@Feliciano2018], but some of the AFS studied here have particularly high stocks in their biomass [compared to the average of 74\ Mg\ C\ ha$^{-1}$ in @Ma2020]. The AFS with the highest carbon stocks, forest gardens, had stocks close to those of old-growth tropical forests [270\ Mg\ C\ ha$^{-1}$; @Lal2005]. These high carbon stocks could be explained by the fact that forest gardens have generally grown for more than 50 years, with little human intervention (Table S1). In terms of structure and carbon stocks, forest gardens should therefore resemble old secondary forests, which have been shown to recover more than 60% of the AGC stocks and almost all the SOC stocks of an old forest [@Poorter2021]. In addition, FGs are generally located close to villages, where soils are enriched by organic waste and animal manure, which can enhance tree growth [@Madejon2016].

Although agroforestry has been identified as one of the possible levers to increase carbon storage in the Timorese agricultural landscape [@UNFCCC2022;@Paudel2022], to our knowledge no study has quantified carbon stocks in these systems. Our study could therefore help to better quantify the national contribution of agroforestry to Timor-Leste's climate change mitigation commitments [@UNFCCC2022], and the potential carbon gains from wider adoption of these techniques. Further studies should help to increase the representativeness of the AFS sampled by including a greater diversity of AFS and environmental conditions. Another critical aspect is the quantification of temporal changes in carbon stocks, which could be assessed in future studies either by having chronosequences of plots where the time since installation is known, or by remeasuring plots through time.

## AFS differ in C stored in their biomass

The AFS studied here vary considerably in terms of AGC stocks, in agreement with our hypothesis H1a. This reflects the density and size of trees in the plot, which in turn depends on the importance of trees in the production system. In some systems, such as CF and SP, trees do not contribute to food production but have other functions, such as restoring soil fertility (e.g. during the fallow period) or providing shade for livestock. This more limited role for trees may explain why their total biomass is lower in these systems: in the CF system, the plot is cleared and annual crops are planted in full sun, while in the SP system, the growth of grazing grasses is favored by reducing the number of trees that shade them.
